---
title: "Cascading shocks in the cereals trade network"
date: "12/12/2015"
output: 
    pdf_document:
        fig_caption: yes
bibliography: ref.bib
---

```{r load_packages, include=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(DiagrammeR)
library(knitr)
```

```{r global_options, include=FALSE}
options(digits = 3)
opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)
```

```{r load_funcs}
source("tfs_cascade_funcs.R")
source("tfs_map_funcs.R")
source("res_00_funcs.R")
```

# Methods

## Simulation model

Our model simulates the impact of a "shock" to the supply of some food commodity on the global trade network for that commodity. The shock is initiated as a drop in production in one or more countries and propagates through the network over multiple intertions of the simulation loop, which is summarized in the flow chart below. 

At each iteration, affected countries first tap into their reserves. When reserves are depleted, countries absorb a fraction of the residual shock by reducing domestic consumption, then reduce their trade balance by decreasing exports and increasing exports, with the impact spread to each trade link proportionally to the volume on that link. This effectively propagates the shock to those countries' trade partners. Finally, any shock that could not be propagated is absorbed by domestic consumption, and the cycle is repeated until all shocks have been absorbed (see the Appendix a detailed description of the model). 

To avoid arbitrarily small shocks being propagated, countries will absorb (through consumption) any shock smaller than a fraction $\alpha$ of their current supply; we set $\alpha$ to 0.001% for all our simulations.

```{r variables_table}
kable(data.frame(rbind(
    c("$p$", "Production by country"),
    c("$r$", "Reserves by country"),
    c("$F$", "Trade matrix ($F_{jk}$ = exports from country $j$ to country $k$)"),
    c("$e$", "Exports by country i.e. $\\sum_k F_{jk}$"),
    c("$i$", "Imports by country i.e. $\\sum_k F_{kj}$"),
    c("$c$", "Domestic consumption by country (for any use)"),
    c("$s$", "Net supply, $s = p + i - e$"),
    c("$\\alpha$", "Minimum threshold (as fraction of $s$) for a shock to be propagated"),
    c("$f_c$", "Fraction of residual shock absorbed by $c$ if $r$ is depleted"),
    c("$f_r$", "Fraction of actual reserves that are available to absorb shocks"),
    c("$f_p$", "Magnitude of initial shock as a fraction of the affected country's $p$")
)), caption = "List of variables and parameters", col.names = c("Symbol", "Description"))
```

```{r flow_chart}
#system("dot -Tpng model-flow-chart.gv -o model-flow-chart.png")
system("dot -Tpdf model-flow-chart.gv -o model-flow-chart.pdf")
```

![Simulation model flow chart](model-flow-chart.pdf)

## Data

We used data from the Food and Agricultural Organization of the United Nations' online database (FAOSTAT, faostat3.fao.org) to construct the cereals production and trade data. The production and trade data for individual primary and secondary commodities in the cereals group was converted into kcal equivalents [@fao2001] and aggregated by country and year (*supplementary table to be added later*). We used population data from FAOSTAT to subset the network so that only countries with a population exceeding half a million people during the period 1986-2011. We rectified the data as described in @carr2013 to account for the merging and fracturing of countries between years.

We obtained data on countries' cereals reserves from the Production, Supply and Distribution database of the United States Department of Agriculture's Foreign Agricultural Service (USDA-PSD, apps.fas.usda.gov/psdonline/). Reserves for nine commodities (barley, corn, millet, mixed grain, oats, milled rice, rye, sorghum and wheat) were converted in kcal equivalents and aggregated by country and year, and political entities were rectified to match the FAO data above. Note that since USDA-PSD reports aggregated reserves for the European Union (pre-1998) and EU-25 (1998 and after), we chose to apportion these reserves between EU countries for each year, in proportion to their share of the EU cereals production. 

## Simulation scenarios

As input data for our model, we used the cereals production, reserves and trade averaged over three five-year periods (1994-1998, 2001-2005 and 2007-2011), which we refer to by their median years (1996, 2003 and 2009). The evolution of the cereals trade across these three time periods show a decrease of global reserves and an increase in global trade, both expressed as a fraction of total production.

```{r load_trade_data}
years <- c(1996, 2003, 2009)
trade_dat <- lapply(years, get_trade_data, mov_avg = 2,
                    prod_trade_file = "data/cereals_prod_trade.RData", 
                    stocks_file = "data/cereals_stocks.RData")
names(trade_dat) <- years
trade_st <- bind_rows(lapply(trade_dat, get_trade_stats_sum), .id = "year")
kable(trade_st, 
      col.names = c("year", "$\\sum p$ (kcal)", "$\\sum r$ (kcal)", 
                    "# trade links", "$\\sum F$ (kcal)",  
                    "$\\sum r / \\sum p$", "$\\sum F / \\sum p$"),
      caption = paste("Summary statistics of the cereals trade data",
                      "for each time period considered in our analysis"))
```

To separate the effects of changes in reserves and trade flows in our results, we performed two different scalings of the 1996 and 2003 reserves: in the $r$-scaled version, all countries' reserves are scaled by a common factor so that the ratio of total $r$ to total $p$ matches that of 2009; in the $r/s$-scaled version, the reserves are set individually in each country so that the fraction of $r$ to net supply $s$ matches the 2009 data. Compared with the original data, simulation results from these scaled versions allow us to identify the impact of changes in total reserves or in the distribution of these reserves between countries, respectively.

```{r r_scaling}
Rscale <- trade_st$RP_ratio / trade_st$RP_ratio[trade_st$year == "2009"]
trade_Rscaled <- Map(
    function(dat, scal) {
        dat$R0 <- dat$R0 / scal
        dat
    }, trade_dat, Rscale)

S0 <- lapply(trade_dat, function(x) x$P0 + colSums(x$E0) - rowSums(x$E0))
RSscale <- trade_dat$`2009`$R0 / S0$`2009`
trade_RSscaled <- Map(
    function(dat, s0) {
        dat$R0 <- RSscale[names(s0)] * s0
        dat
    }, trade_dat, S0)
```

We define a model run as a set that includes one simulation starting from each country for each of the three years. In addition to the input data (original, $r$-scaled or $r/s$-scaled version), we set three global parameters by model run: the initial shock magnitude as a fraction of the target country's production ($f_p$), the fraction of reserves that are available to absorb a shock ($f_r$), as well as the fraction of a shock absorbed by consumption after reserves are depleted ($f_c$, as defined in our model above). In practice, $f_r < 1$ may reflect the fact that some of the reported reserves are already committed to trade or that countries are unwilling to let reserves drop below some floor amount. 

```{r run_sims}
get_multiyear_sim_results <- function(trade_data, pfrac, rfrac, cfrac) {
    res <- lapply(trade_data, 
                  function(x) sim_allc(pfrac, x$P0, rfrac*x$R0, x$E0, cfrac))
    st <- lapply(res, get_stats_allc)
    trade_stats <- bind_rows(lapply(trade_data, get_trade_stats_by_cty),
                             .id = "year") 
    sim_stats <- bind_rows(lapply(st, get_sim_stats_by_cty))
    cbind(trade_stats, sim_stats)
}

#stats_by_cty <- bind_rows(
#    "02_05"    = get_multiyear_sim_results(trade_dat, 0.2, 0.5, 0.01),
#    "Rscaled"  = get_multiyear_sim_results(trade_Rscaled, 0.2, 0.5, 0.01),
#    "RSscaled" = get_multiyear_sim_results(trade_RSscaled, 0.2, 0.5, 0.01),
#    "01_025"   = get_multiyear_sim_results(trade_dat, 0.1, 0.25, 0.01),
#    "03_1"     = get_multiyear_sim_results(trade_dat, 0.3, 1, 0.01),
#    "015_05"    = get_multiyear_sim_results(trade_dat, 0.15, 0.5, 0.01),   
#    "02_05_005"= get_multiyear_sim_results(trade_dat, 0.2, 0.5, 0.05),
#    "02_05_01" = get_multiyear_sim_results(trade_dat, 0.2, 0.5, 0.1),    
#    .id = "params")
```


...

# Results




# References

FAO (2001). Food balance sheets: A handbook.

Carr, J. A., D’Odorico, P., Laio, F., & Ridolfi, L. (2013). Recent history and geography of virtual water trade. *PLoS ONE*, **8**(2): e55825.



# Appendix: Simulation model details

In our model, we define the annual net supply (or simply "supply") $s_j$ of a food commodity in country $j$ as:
$$s_j = p_j + i_j - e_j \, ,$$ 
where $p$ is production, $i$ is imports and $e$ is exports. The difference between this supply and domestic consumption $c$ must result in a change in the country's reserves $r$, so that: 
$$s_j = c_j + \Delta r_j \, .$$ 
Note that this convention differs from the FAO's definition of domestic supply, which includes a term representing transfer *from* the reserves, whereas a positive $\Delta r$ in our case represents a transfer *to* the reserves. 

We initialize the model with realistic production, trade and reserves data and assume that the initial consumption equals supply (i.e. $\Delta r = 0$ before any shock). The trade data is represented by a matrix $F$ where $F_{jk}$ is the volume of exports from country $j$ to country $k$. 

In the first iteration of the model ($t$ = 1), a supply shock is initiated from a drop in production:
$$ \Delta s'_{1,j} = \Delta p_{j} \, ,$$
where $\Delta p_{j} = 0$ for all but one or a few countries. In the following iterations, $\Delta s'$ will be due to decrease in imports or increase in exports as the shock is propagated through trade. The prime ($'$) notation is used to distinguish different temporary values of the shock within a single iteration. 

At the start of a new iteration, all countries receiving a new shock (form any source) first attempt to absorb it through their reserves:
$$ \Delta r_{t,j} = \max \left( \Delta s'_{t,j}, - r_{t,j} \right) \, . $$ 

Any country that has depleted its reserves (or had none to begin with) will block any increase in its outgoing trade and is thus removed from the set $\mathcal{U}$ (which initially includes all countries).
$$ \mathcal{U} = \mathcal{U} \setminus 
    \left\{ j \mid \Delta s'_{t,j} < - r_{t,j} \right\} \, . $$
    
At this point, countries left with a residual shock absorb a fraction of it ($f_c$) by reducing domestic consumption, leaving an amount $\Delta s''_{t,j}$ that can be transferred to other countries through trade:                 
$$ \Delta s''_{t,j} = \left( \Delta s'_{t,j} - \Delta r_{t,j} \right)
                      \left( 1 - f_c \right) \, . $$ 
To avoid spending a large fraction of the simulation passing tiny residual shocks, we define a fractional threshold $\alpha$ so that countries simply absorb their residual shock when $\lvert \Delta s''_{t,j} \rvert < \alpha \, s_{t,j}$. 

A country can transfer a shock to its trade partners by either reducing its exports or increasing its imports, i.e. by decreasing its balance of trade $b$. We set an upper bound on the magnitude of this change to be equal to the total volume $v$ on all the country's trade links, excluding any "blocked" incoming links:
$$ v_{t,j} = \sum_k F_{t,jk} + \sum_k F_{t,kj} \, \mathbf{1}_{\mathcal{U}}(k)$$ and
$$ \Delta b_{t,j} = \max \left( \Delta s''_{t,j}, - v_{t,j} \right) \, ,$$ 
where $\mathbf{1}_{\mathcal{U}}(k)$ equals 1 if $k$ is in $\mathcal{U}$ and 0 otherwise.  

Each country then changes the trade volume on each unblocked link (- if outgoing, + if incoming) proportionally to the current volume on that link. From this, we can calculate the change of any element of the trade matrix as:
$$ \Delta F_{t,jk} = \left[ \left( \frac{\Delta b_{t,j}}{ v_{t,j}} \right)  
                          - \mathbf{1}_{\mathcal{U}}(j) 
                            \left( \frac{\Delta b_{t,k}}{ v_{t,k}} \right)
                            \right] F_{t,jk} \, .$$

Finally, any portion of the initial shock not absorbed by reserves or passed through trade results in a decrease in consumption:
$$ \Delta c_{t,j} = \Delta s'_{t,j} - \Delta r_{t,j} - \Delta b_{t,j} \, .$$

To start the next iteration, we update the vectors $r$, $c$ and $s$ as well as the trade matrix $F$ based on the computed $\Delta$. Note that the actual change in $s$ for a given country only depends on the fraction of the shock that was absorbed internally:
$$ \Delta s_{t,j} = \Delta r_{t,j} + \Delta c{t,j} \, .$$
The new shock received by each country at the next iteration equal to any increase in exports or decrease in imports due to other countries' actions at this iteration, i.e.:
$$ \Delta s'_{t+1,j} = \sum_k \Delta F_{t,kj} - \sum_k \Delta F_{t,jk} 
                      + \Delta b_{t,j} \, ,$$
where the last term serves to remove the effect of the country's own actions. When $\Delta s'_{t+1, j} = 0$ for all $j$, all shocks have been absorbed internally and the simulation ends.