---
title: "Supplementary Data - Reserves and trade jointly determine exposure to food supply shocks"
output:
  pdf_document:
    fig_caption: yes
---

```{r load_packages, include=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2) 
library(knitr)
library(stringr)
library(xtable) # LaTeX table formatting 
library(cowplot) # for multi-panel plots
library(FAOSTAT) # to acquire FAOSTAT data
source("cereals-network-funcs.R")
```

```{r global_options, include=FALSE}
options(digits = 3, xtable.comment = FALSE)
opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)
```


# Simulation model details

In our model, we define the annual net supply (or simply "supply") $S_j$ of a food commodity in country $j$ as:
$$S_j = P_j + I_j - E_j \, ,$$ 
where $P$ is production, $I$ is imports and $E$ is exports. The difference between this supply and domestic consumption $C$ must result in a change in the country's reserves $R$, so that: 
$$S_j = C_j + \Delta R_j \, .$$ 
Note that this convention differs from the FAO's definition of domestic supply, which includes a term representing transfer *from* the reserves, whereas a positive $\Delta R$ in our case represents a transfer *to* the reserves. 

We initialize the model with realistic production, trade and reserves data and assume that the initial consumption equals supply, i.e. $\Delta R = 0$ before any shock. The trade data is represented by a matrix $F$ where $F_{jk}$ is the volume of exports from country $j$ to country $k$. 

In the first iteration of the model ($t$ = 1), a supply shock is initiated from a drop in production:
$$ \Delta S'_{1,j} = \Delta P_{j} \, ,$$
where $\Delta P_{j} = 0$ for all but one or a few countries. In the following iterations, $\Delta S'$ is caused by a decrease in imports or an increase in exports as the shock is propagated through trade. The prime ($'$) notation is used to distinguish different temporary values of the shock within a single iteration. 

At the start of an iteration, all countries receiving a new shock (from any source) first attempt to absorb it through their reserves:
$$ \Delta R_{t,j} = \max \left( \Delta S'_{t,j}, -R_{t,j} \right) \, . $$ 

Any country that has depleted its reserves (or had none to begin with) will block any increase in its outgoing trade and is thus removed from the set $\mathcal{U}$, which initially includes all countries.
$$ \mathcal{U} = \mathcal{U} \setminus 
    \left\{ j \mid \Delta S'_{t,j} < -R_{t,j} \right\} \, . $$
    
At this point, countries left with a residual shock absorb a fraction of it ($f_c$) by reducing domestic consumption, leaving an amount $\Delta S''_{t,j}$ that can be transferred to other countries through trade:                 
$$ \Delta S''_{t,j} = \left( \Delta S'_{t,j} - \Delta R_{t,j} \right)
                      \left( 1 - f_c \right) \, . $$ 
To avoid spending a large fraction of the simulation time passing tiny residual shocks, we define a fractional threshold $\alpha$ so that countries simply absorb their residual shock when $\lvert \Delta S''_{t,j} \rvert < \alpha \, S_{t,j}$. 

A country can transfer a shock to its trade partners by either reducing its exports or increasing its imports, i.e. by decreasing its balance of trade $B$. We set an upper bound on the magnitude of this change to be equal to the total volume $V$ on all the country's trade links, excluding any "blocked" incoming links:
$$ V_{t,j} = \sum_k F_{t,jk} + \sum_k F_{t,kj} \, \mathbf{1}_{\mathcal{U}}(k)$$ and
$$ \Delta B_{t,j} = \max \left( \Delta S''_{t,j}, -V_{t,j} \right) \, ,$$ 
where $\mathbf{1}_{\mathcal{U}}(k)$ equals 1 if $k$ is in $\mathcal{U}$ and 0 otherwise.  

Each country then changes the trade volume on each unblocked link (- if outgoing, + if incoming) proportionally to the current volume on that link. From this, we can calculate the change of any element of the trade matrix as:
$$ \Delta F_{t,jk} = \left[ \left( \frac{\Delta B_{t,j}}{ V_{t,j}} \right)  
                          - \mathbf{1}_{\mathcal{U}}(j) 
                            \left( \frac{\Delta B_{t,k}}{ V_{t,k}} \right)
                            \right] F_{t,jk} \, .$$

Finally, any portion of the initial shock not absorbed by reserves or passed through trade results in a decrease in consumption:
$$ \Delta C_{t,j} = \Delta S'_{t,j} - \Delta R_{t,j} - \Delta B_{t,j} \, .$$

To start the next iteration, we update the vectors $R$, $C$ and $S$ as well as the trade matrix $F$ based on the computed $\Delta$. Note that the actual change in $S$ for a given country only depends on the fraction of the shock that was absorbed internally:
$$ \Delta S_{t,j} = \Delta R_{t,j} + \Delta C_{t,j} \, .$$
The new shock received by each country at the next iteration equals the sum of all increases in exports or decreases in imports due to other countries' actions at this iteration, i.e.:
$$ \Delta S'_{t+1,j} = \sum_k \Delta F_{t,kj} - \sum_k \Delta F_{t,jk} 
                      + \Delta B_{t,j} \, ,$$
where the last term serves to remove the effect of the country's own actions. When $\Delta S'_{t+1, j} = 0$ for all $j$, all shocks have been absorbed internally and the simulation ends.


# Sensitivity analysis for global simulation parameters

Our model includes three adjustable global simulation parameters: the fraction of domestic production lost in the initial shock ($f_p$), the fraction of reserves available to buffer a shock ($f_r$) and the fraction of a shock absorbed by domestic consumption before it is passed on to trade partners ($f_c$). In Table \ref{tab:pars_stat}, we show how changes in these parameters affect the simulation results, using the most recent (2009) data version. For a more detailed description of the output statistics, see the main text of the article.

Between the first two rows of Table \ref{tab:pars_stat}, both $f_p$ and $f_r$ are doubled so that the ratio of shock intensity to initial reserves remains constant. This leaves most impact metrics unaffected, except for the mean relative change in consumption ($\Delta c_{rel}$), which nearly doubles. In this scenario, even if countries have larger available reserves, their trade flows remain the same and may limit the amount of residual shock that can be passed along, resulting in a greater impact on consumption. The same effect would explain the decrease in evenness $J$ between the two cases. Predictably, raising $f_c$ (third and fouth rows) monotonically increases the magnitude of $\Delta c_{rel}$ while leaving other statistics unchanged.

```{r tab_pars_stat, results = "asis"}
# Calculate summary statistics for model runs using different fp, fr, fc
# Print output table
stats_by_cty <- readRDS("stats_by_cty.RData")

Cfrac <- c("02_05", "02_05_005", "02_05_01") # effect of fc parameter
PR <- c("01_025", "015_05", "02_05", "03_1")  # effect of fp, fr parameters
hits_summ <- stats_by_cty %>%
    filter(params %in% c(Cfrac, PR), year == 2009) %>%
    group_by(params) %>%
    summarise(fp = mean(fp), fr = mean(fr), fc = mean(fc),
              nshocks = sum(self_dC_scl < 0), 
              hits_m = mean(hits), hits_sd = sd(hits),
              mean_tot_dS = mean(tot_dS_scl), sd_tot_dS = sd(tot_dS_scl),
              mean_even = mean(even), sd_even = sd(even),
              hitsC = mean(hitsC), mean_tot_dC = mean(tot_dC_scl)) %>%
    select(-params) %>% arrange(desc(nshocks))

titlerow <- list(pos = list(0, 0),
    command = c(paste("$f_p$ & $f_r$ & $f_c$ & $N_s$ &",
                      "\\multicolumn{2}{c}{$N_{h}$} &", 
                      "\\multicolumn{2}{c}{$\\Delta s_{rel}$} &",
                      "\\multicolumn{2}{c}{$J$} & $N_{hc}$ &",
                      "$\\Delta c_{rel}$ \\\\\n"),
                paste("& & & & mean & s.d. & mean & s.d. &",
                      "mean & s.d. & mean & mean \\\\\n")))
pars_stat <- xtable(hits_summ, label = "tab:pars_stat",
    caption = paste("Effect of global parameters on simulation outputs using",
                    "the 2009 data. Each row aggregates results from a set of",
                    "simulations, each with a shock originating",
                    "in a different country. $N_s$ is the number of simulations where", 
                    "a shock is passed; $N_h$ (resp., $N_{hc}$) is",
                    "the number of times a country's supply (resp., consumption)",
                    "are affected across simulations; $\\Delta s_{rel}$",
                    "(resp., $\\Delta c_{rel}$) is a relative measure of the total",
                    "change in a country's supply (resp., consumption) across simulations;",
                    "$J$ is the evenness of impact between simulations. Means and",
                    "standard deviations are calculated across affected countries.",
                    "\\newline"),
    align = "lrrr|rrrrrrrrr")
print(pars_stat, add.to.row = titlerow, caption.placement = "top",
      include.rownames = FALSE, include.colnames = FALSE, 
      hline.after = c(-1, 0, 2, 4, 6))
```

Compared to the first two cases, the last two rows in Table \ref{tab:pars_stat} represent a 50% increase in shock intensity and a doubling of reserves. As more countries can absorb a shock through their reserves, the number of simulations where a shock is passed ($N_s$), as well as the mean number of hits ($N_{h}$ and $N_{hc}$) by country all decrease. Since a larger portion of the shock is absorbed by the initial target, shocks become less spread out and the evenness $J$ decreases.

```{r prep_dc09}
stats_by_cty <- readRDS("stats_by_cty.RData")

dC_09 <- stats_by_cty %>%
    filter(year == "2009", !(params %in% c("Rscaled", "RSscaled"))) %>%
    mutate(ext_dC_scl = tot_dC_scl - self_dC_scl,
           ext_dC_S0 = tot_dC_S0 - self_dC_S0) %>%
    group_by(cty)


dCself09 <- filter(dC_09, min(self_dC_S0) < -0.05)
    
cty_order <- filter(dCself09, params == "02_05") %>%
    ungroup() %>% arrange(self_dC_S0)

dCself09$cty <- factor(dCself09$cty, levels = cty_order$cty)
dCself09$params <- as.factor(dCself09$params)

dCext09 <- filter(dC_09, min(ext_dC_S0) < -0.05)

cty_order <- filter(dCext09, params == "02_05") %>%
    ungroup() %>% arrange(ext_dC_S0)

dCext09$cty <- factor(dCext09$cty, levels = cty_order$cty)
dCext09$params <- as.factor(dCext09$params)
```

```{r fig_dc09, fig.width = 6, fig.height = 8, fig.cap = "\\label{fig:dc09} Decrease in consumption ($\\Delta C$) as a fraction of initial net supply ($S_0$) from (a) internal shocks (i.e. shock was initiated at the target country) and (b) external shocks (initiated at other countries), for varying settings of the global parameters $f_p$, $f_r$ and $f_c$. The trade network was initialized with 2009 data. The countries shown in each plot (identified by their ISO alpha-3 codes) include all those experiencing a 5% decrease in consumption for at least one parameter set."}

p1 <- ggplot(data = dCself09, aes(x = cty, y = self_dC_S0, 
                                  color = params, group = params)) +
    labs(x = "country", y = expression(Delta * C[internal] / S[0])) +
    geom_point() + geom_line() +
    scale_color_brewer(type = "qual") +
    theme(axis.text.x = element_text(angle = 45, size = 8),
          axis.text.y = element_text(size = 8),
          legend.position = "none")

p2 <- ggplot(data = dCext09, aes(x = cty, y = ext_dC_S0, 
                                 color = params, group = params)) +
    labs(x = "country", y = expression(Delta * C[external] / S[0]), color = "") +
    geom_point() + geom_line() +
    scale_color_brewer(type = "qual", 
        labels = c(expression(paste(f[p] == 0.1, ", ", f[r] == 0.25, ", ", f[c] == 0.01)),
                   expression(paste(f[p] == 0.15,", ", f[r] == 0.5, ", ", f[c] == 0.01)),
                   expression(paste(f[p] == 0.2, ", ", f[r] == 0.5, ", ", f[c] == 0.01)),
                   expression(paste(f[p] == 0.2, ", ", f[r] == 0.5, ", ", f[c] == 0.05)),
                   expression(paste(f[p] == 0.2, ", ", f[r] == 0.5, ", ", f[c] == 0.1)),
                   expression(paste(f[p] == 0.3, ", ", f[r] == 1, ", ", f[c] == 0.01)))
        ) +
    theme(axis.text.x = element_text(angle = 45, size = 8),
          axis.text.y = element_text(size = 8),
          legend.position = "bottom",
          legend.text = element_text(size = 8))

plot_grid(p1, p2, labels = c("a", "b"), ncol = 1)
```

Figure \ref{fig:dc09} shows the effect of global simulation parameters on the decrease in domestic consumption for the countries most affected by internal and external production shocks. Countries are ordered along the $x$-axis based on the magnitude of impact in the base scenario ($f_p$ = 0.2, $f_r$ = 0.5 and $f_c$ = 0.01). While the impact on $\Delta C$ increases with the magnitude of the initial shock (i.e. increasing $f_p$), the set of most affected countries is fairly consistent. Note that the exact ordering of these countries across parameter sets may vary based on their specific position in the trade network. For example, since a proportion $f_c$ of each shock received is absorbed by consumption, trade hubs such as Singapore (SGP) and Qatar (QAT) that receive many external shocks are most affected by a change in this parameter (Figure \ref{fig:dc09}b).


# Additional tables and figures

```{r tab:crop_list, results = "asis"}
crop_list <- read.delim("data/crop_list.tsv", stringsAsFactors = FALSE) %>%
    filter(Group == "1 - CEREALS AND CEREAL PRODUCTS") %>%
    select(cropid, cropname, ProductionL, kcal.ton) %>%
    mutate(ProductionL = as.logical(ProductionL), 
           kcal.ton = as.numeric(kcal.ton)*1E-6)

colnames(crop_list) <- c("FAOSTAT item code", "Crop name", 
                         "Primary?", "Conversion factor ($10^6$ kcal/ton)")

print(xtable(crop_list, label = "tab:crop_list", digits = 2,
             caption = paste("List of cereal commodities from FAOSTAT and their calorie",
                             "content (based on FAO 2001). The aggregate production",
                             "data includes only primary commodities whereas the",
                             "trade data includes all commodities. Trade figures for",
                             "total rice - milled equivalent (code 30) were treated as", 
                             "milled rice (code 31). \\newline")), 
      caption.placement = "top", include.rownames = FALSE,
      sanitize.colnames.function = function(x) {x}, scalebox = 0.8)
```

```{r fbs_prep, results = "hide"}
country_list <- read.csv("data/country_list.csv", stringsAsFactors = FALSE)
yr_range <- 1994:2011

# Get food balance and pop. data from FAOSTAT
#  domain BC and item 2905 correspond to the commodity balance for cereals (excl. beer)
#  domain OA and item 3010 correspond to the annual total population by country
foodbal <- getFAOtoSYB(name = c("dR", "Waste", "Proc", "Food", "Other", "S", 
                                 "P", "Feed", "Seed", "I", "E", "pop"),
                       elementCode = c("5071", "5120", "5130", "5141", "5153",
                         "5300", "5510", "5520", "5525", "5610", "5910", "511"),
                       domainCode = c(rep("BC", 11), "OA"),
                       itemCode = c(rep("2905", 11), "3010"),
                       yearRange = yr_range, countrySet = country_list$FAOST_CODE,
                       useCHMT = FALSE, outputFormat = "long")
foodbal <- select(foodbal$entity, -domainCode, -itemCode, -elementCode) %>%
    spread(key = name, value = Value)

# Only keep countries with maxpop >500k and P > 0 each year
has_prod <- group_by(foodbal, FAOST_CODE) %>%
    summarise(has_prod = all(P > 0))
cty_keep <- intersect(country_list$FAOST_CODE[country_list$gt.500k],
                      has_prod$FAOST_CODE[has_prod$has_prod])
foodbal <- filter(foodbal, FAOST_CODE %in% cty_keep)

# Get values per capita and annual differences
#  Note the change of sign for dR
cereals_percap <- foodbal %>%
    transmute(cty = FAOST_CODE, year = Year, P = P / pop, dR = -dR / pop, 
              E = E / pop, I = I / pop, S = S / pop, food = Food / pop,
              feed = Feed / pop, seed = Seed / pop, proc = Proc / pop,
              waste = Waste / pop, other = Other / pop) %>%
    group_by(cty) %>% arrange(cty, year)

cereals_percap <- mutate(cereals_percap, dP = P - lag(P), ddR = dR - lag(dR), 
                         dE = E - lag(E), dI = I - lag(I), dS = S - lag(S),
                         dfood = food - lag(food), dfeed = feed - lag(feed),
                         dseed = seed - lag(seed), dproc = proc - lag(proc), 
                         dwaste = waste - lag(waste), dother = other - lag(other))
```

```{r tab:dP_reg, results = "asis"}
# Shock when dP/P < -0.1
cereals_shock <- mutate(cereals_percap, dPrel = dP / lag(P)) %>%
    filter(dPrel < -0.1)

# Get regression coefs and R^2 values for all food balance items vs. dP
get_dP_reg <- function(var, df) {
    linm <- lm(df[[var]] ~ df[["dP"]])
    c(b = coefficients(linm)[2], 
      rsq = summary(linm)$adj.r.squared,
      sd = sd(linm$residuals))
}
vars <- c("ddR", "dE", "dI", "dfood", "dfeed",
          "dseed", "dproc", "dwaste", "dother")
dP_fit <- cbind(
    t(vapply(vars, get_dP_reg, df = cereals_percap, rep(0, 3))),
    t(shock_fit <- vapply(vars, get_dP_reg, df = cereals_shock, rep(0, 3)))
)

rownames(dP_fit) <- c("$\\Delta(\\Delta R)$", "$\\Delta E$", "$\\Delta I$", 
                      "$\\Delta C_{food}$", "$\\Delta C_{feed}$", "$\\Delta C_{seed}$",
                      "$\\Delta C_{process}$", "$\\Delta C_{waste}$", "$\\Delta C_{other}$")

titlerow <- list(pos = list(0, 0),
    command = c(paste("& \\multicolumn{3}{c}{All data} &",
                      "\\multicolumn{3}{c}{$\\Delta P < -0.1$} \\\\\n"),
                paste("& $\\beta$ & $R^2$ & $\\sigma_{res}$ &",
                      "$\\beta$ & $R^2$ & $\\sigma_{res}$ \\\\\n")))
dP_reg <- xtable(as.data.frame(dP_fit), label = "tab:dP_reg", align = c("l|rrr|rrr"),
    caption = paste("Regression slope ($\\beta$), adjusted $R^2$, and residual standard",
                    "deviation ($\\sigma_{res}$) for the effect of the interannual $\\Delta P$",
                    "on each other component of the FAO food balance sheets (FBS)",
                    "for cereals and cereal products, extracted by country and year and",
                    "expressed on a per capita basis. Note that $\\Delta R$ is a flow \\emph{to}", 
                    "reserves (opposite of the FBS sign convention) and $\\Delta(\\Delta R)$ is",
                    "the interannual change in that flow. We used the FBS data from 1994 to 2011",
                    "for the 153 countries that reported cereals production for each year in this",
                    "period and reached a population of at least 500,000. The right-hand side",
                    "results are based on a subset of the country-years where per capita",
                    "production fell by over 10\\%. \\newline"))

print(dP_reg, add.to.row = titlerow, caption.placement = "top", 
      include.colnames = FALSE, hline.after = c(-1, 0, 9),
      sanitize.rownames.function = function(x) {x})

```

```{r stats_vs_S, fig.width = 4, fig.height = 8, fig.cap = "\\label{fig:stats_vs_S} (a) $R$/$S$ ratio and (b) network degree (total number of incoming and outgoing links) as a function of $\\log S$ for the 2009 network. Labels identify (via their ISO alpha-3 codes) the top 10% exporters, responsible for ~86% of total exports."}

# Get trade data for 2009
trade_dat <- get_trade_data(2009, prod_trade_file = "data/cereals_prod_trade.RData", 
                            stocks_file = "data/cereals_stocks.RData", mov_avg = 2)
trade_by_cty <- get_trade_stats_by_cty(trade_dat)

# Pick 10% top exporters
trade_10pct <- filter(trade_by_cty, percent_rank(E0_S0 * S0) >= 0.9)

p1 <- ggplot(data = trade_by_cty, aes(x = log(S0), y = R0_S0)) +
    labs(x = "", y = "R/S") +
    geom_point(alpha = 0.5) + 
    geom_point(data = trade_10pct) +
    geom_smooth(method = "lm", se = FALSE) +
    geom_text(data = trade_10pct, aes(label = cty), size = 3, 
              nudge_x = 0.6, nudge_y = 0.012)

p2 <- ggplot(data = trade_by_cty, aes(x = log(S0), y = outdeg + indeg)) +
    labs(x = "log S", y = "degree") +
    geom_point(alpha = 0.5) + 
    geom_point(data = trade_10pct) + 
    geom_smooth(method = "lm", se = FALSE) +
    geom_text(data = trade_10pct, aes(label = cty), size = 3,  
              nudge_x = 0.6, nudge_y = 5)

plot_grid(p1, p2, labels = c("a", "b"), ncol = 1)
```